---
title: "10. Random forest for key molecular investigate"
author: "Tao, Yang"
date: "2024-03-06"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---
### Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(Seurat)
library(SeuratObject)
library(randomForest)
library(ggplot2)
library(ggExtra)
library(ggpmisc)
library(ggpubr)
library(data.table)
library(forcats)
library(RColorBrewer)
```

### Load data
```{r}
zf10_ligands <- readRDS(here::here("output","temp","zf10_ligands.rds")) %>% as.data.frame()
zf10_TF <- readRDS(here::here("output","temp","zf10_TF.rds")) %>% as.data.frame()
```

```{r}
input <- zf10_ligands

input$AP_index <- input$rowname
input_ <- input[2:length(colnames(input))]
input_$AP_index <- as.numeric(input_$AP_index)
```

### 70% data to train random forest model, 30% data for test
```{r}
set.seed(120)
train <- sample(nrow(input_), nrow(input_)*0.7)
colnames(input_) <- gsub(":", "_", colnames(input_)) 
colnames(input_) <- gsub("-", "_", colnames(input_)) 
AP_train <- input_[train, ]
AP_test <- input_[-train, ]

set.seed(123)
AP_train.forest <- randomForest(AP_index~., data = AP_train, importance = TRUE)
```

### Use training data to check the model accuracy
```{r}
train_test <- predict(AP_train.forest, AP_train)
train_test_dataframe <- as.data.frame(train_test)
train_test_dataframe$index <- rownames(train_test_dataframe)
AP_train_dataframe <- AP_train %>% rownames_to_column("index")

train_compare <- left_join(train_test_dataframe,AP_train_dataframe)

g <- ggplot(train_compare, aes(AP_index, train_test)) + 
  geom_point() + 
  geom_smooth(method="lm", se=F) +
  geom_abline(slope = 1,intercept = 0,lty="dashed") +
  stat_poly_eq(
    aes(label =paste( ..adj.rr.label.., sep = '~~')),
    formula = y ~ x,  parse = TRUE,
    family="serif",
    size = 6.4,
    color="black",
    label.x = 0.1, 
    label.y = 1)

g1 <- ggMarginal(g, type = "histogram", fill="transparent")
```

### Use test data to check model performance
```{r}
predict_test <- predict(AP_train.forest, AP_test)
predict_test_dataframe <- as.data.frame(predict_test)
predict_test_dataframe$index <- rownames(predict_test_dataframe)
AP_test_dataframe <- AP_test %>% rownames_to_column("index")
test_compare <- left_join(AP_test_dataframe,predict_test_dataframe)

g <- ggplot(test_compare, aes(AP_index, predict_test)) + 
  geom_point() + 
  geom_smooth(method="lm", se=F) +
  geom_abline(slope = 1,intercept = 0,lty="dashed") +
  stat_poly_eq(
    aes(label =paste( ..adj.rr.label.., sep = '~~')),
    formula = y ~ x,  parse = TRUE,
    family="serif",
    size = 6.4,
    color="black",
    label.x = 0.1, 
    label.y = 1)

g2 <- ggMarginal(g, type = "histogram", fill="transparent")
ggarrange(g1, g2, ncol = 2)
```

### Access the factor importance to AP position
```{r}
importance_AP <- AP_train.forest$importance
importance_plot <- tibble(var = rownames(importance_AP), 
                          IncMSE = importance_AP[,1],
                          IncNodePurity = importance_AP[,2])
DT <- setDT(importance_plot)

importance_plot$IncMSE <- importance_plot$IncMSE %>% as.numeric()


p1 <- importance_plot %>% 
  arrange(importance_plot$IncMSE) %>% 
  mutate(var=factor(var, levels=var)) %>%
  tail(10) %>%
  ggplot(aes(x=var, y=IncMSE)) +
  geom_segment( aes(x=var, xend=var, y=0, yend=IncMSE), color="skyblue") +
  geom_point( color="blue", size=5, alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

p2 <- importance_plot %>% 
  arrange(importance_plot$IncNodePurity) %>% 
  mutate(var=factor(var, levels=var)) %>%
  tail(10) %>%
  ggplot(aes(x=var, y=IncNodePurity)) +
  geom_segment( aes(x=var, xend=var, y=0, yend=IncNodePurity), color="skyblue") +
  geom_point( color="blue", size=5, alpha=0.6) +
  theme_light() +
  coord_flip() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.border = element_blank(),
    axis.ticks.y = element_blank()
  )

ggarrange(p1, p2, ncol = 2)
```

### Cross validation to evaluate the number of variants
```{r}
set.seed(111)
AP_train.cv <- replicate(5, rfcv(AP_train[-ncol(AP_train)], AP_train$AP_index, cv.fold = 5, step = 0.8), simplify = FALSE)

AP_train.cv <- data.frame(sapply(AP_train.cv, '[[', 'error.cv'))
AP_train.cv$vars <- rownames(AP_train.cv)
AP_train.cv <- reshape2::melt(AP_train.cv, id = 'vars')
AP_train.cv$vars <- as.numeric(as.character(AP_train.cv$vars))

AP_train.cv.mean <- aggregate(AP_train.cv$value, by = list(AP_train.cv$vars), FUN = mean)
AP_train.cv.mean

ggplot(AP_train.cv.mean, aes(Group.1, x)) +
  geom_point(size=2) +
  geom_line(size=1,col="blue") +
  labs(title = '',x = 'Number of vars', y = 'Cross-validation error') + 
  xlim(0,30)
```

### Visualize the top 6 variant distribution on AP axis
```{r}
importance_AP <- importance_plot[order(importance_plot$IncNodePurity, decreasing = TRUE), ]

importance_AP.select <- importance_AP[1:6, ]
vars <- c(pull(importance_AP.select, var), 'AP_index')
AP.select <- input_[,vars]
AP.select <- reshape2::melt(AP.select, id = 'AP_index')

ggplot(AP.select, aes(x = AP_index, y = value)) +
  geom_point(size=1,col="gray") +
  geom_smooth(size=2) +
  facet_wrap(~variable, ncol = 2, scale = 'free_y') +
  labs(title = '',x = 'AP_index', y = 'Relative abundance') +
  theme_bw() + 
  theme(legend.position = "none", panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))

```

### Use a slide window to sum the expression along AP axis
```{r}
s <- 0.1
df <- matrix(NA, nrow = 16514, ncol = 0)

for (i in seq(0,0.9,s)){
  if (count(between(zf10_linearizeAP$normalized_AP,i,i+s)) > 0) {
  subset <- zf10_linearizeAP[,between(zf10_linearizeAP$normalized_AP,i,i+s)]
  sce <- SingleCellExperiment(assays = list(counts = subset@assays$RNA@counts), 
                              colData = subset@meta.data)
  groups <- colData(sce)[, c("orig.ident")]
  bulk <- aggregate.Matrix(t(counts(sce)),groupings = groups, fun = "sum") %>%
  t() %>%
  as.data.frame()
  colnames(bulk) <- i+s/2
  df <- df %>% cbind(bulk)
  }
}

df_norm <- df/do.call(pmax, df)
```

### Heatmap showing the distribution of morphogens and TFs along AP axis
```{r}
df_ligands <- df_norm[c("fgf3","aldh1a2","gdf3","fgf8a","wnt8b","bmp4","wnt5b","jag2b","bmp7a","wnt11"),] %>% t() %>% as.data.frame() %>% 
  rownames_to_column("AP_position")
  
df_ligands <- df_norm[c("fgf3","jag2b","wnt8b","wnt11","aldh1a2","fgf8a","wnt5b","gdf3","bmp4","bmp7a"),] %>% t() %>% as.data.frame() %>% 
  rownames_to_column("AP_position")

pheatmap::pheatmap(
  df_ligands[2:8,] %>% dplyr::select(-AP_position) %>% t(),
  color = colorRampPalette(c("white","#925E9FFF"))(256),
  labels_row = rownames(df_ligands %>% dplyr::select(-AP_position) %>% t()),
  labels_col = df_ligands$AP_position,
  border_color = "white", 
  cluster_rows = F,
  show_colnames = F,
  cutree_cols = 3, 
  cluster_cols = FALSE, number_color = "black",
  cellwidth = 20, cellheight = 10,fontsize = 9,
  #annotation_row = rownames(X),
  main = "Heatmap zf10 key morphogens"
  )


df_TF <- df_norm[c("rx3","vox","six7","gsc","her6","hoxb7a","sox19a","cdx4","irx7","sox3"),] %>% t() %>% as.data.frame() %>% 
  rownames_to_column("AP_position")

df_TF <- df_norm[c("gsc","rx3","six7","her6","sox19a","sox3","irx7","hoxb7a","cdx4","vox"),] %>% t() %>% as.data.frame() %>% 
  rownames_to_column("AP_position")

heatmap <- pheatmap::pheatmap(
  df_TF[2:8,] %>% dplyr::select(-AP_position) %>% t(),
  color = colorRampPalette(c("white","#8bc34a"))(256),
  labels_row = rownames(df_TF %>% dplyr::select(-AP_position) %>% t()),
  labels_col = df_TF$AP_position,
  border_color = "white", 
  cluster_rows = F,
  show_colnames = F,
  cutree_cols = 3, 
  cluster_cols = FALSE, number_color = "black",
  cellwidth = 20, cellheight = 10,fontsize = 9,
  #annotation_row = rownames(X),
  main = "Heatmap zf10 key TFs",
  clustering_callback = callback
  )
```
